---
title: "Tests for Categorical Variables"
author: "Mark Dunning. Original materials by Materials by Deepak Parashar, Sarah Dawson and Sarah Vowler"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: 
  slidy_presentation:
    footer: 'This work is licensed under the Creative Commons Attribution-ShareAlike
      2.0. COURSE HOME: tiny.cc/crukStats'
  ioslides_presentation: default
  beamer_presentation: default
css: mystyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA,echo=FALSE,message=FALSE,warning = FALSE)
```

# Tests for categorical variables

## Associations between categorical variables

- All about frequencies!
- Row x Column table (2 x 2 simplest)
- Categorical data
- Look for association (relationship) between row variable and column variable

```{r}
library(ascii)
library(knitr)
df <- data.frame(Tumour.Did.Not.Shrink = c(44,24), Tumour.Did.Shrink = c(40,16))
rownames(df) <- c("Treatment","Placebo")
print(ascii(df), type="rest")
```

## Chi-square test

- E.g. Research question: A trial to assess the effectiveness of a new treatment versus a placebo in reducing tumour size in patients with ovarian cancer.

```{r}
print(ascii(df),type="rest")
```

- Is there an association between treatment group and tumour shrinkage
- Null hypothesis, $H_0$: No association
- Alternative hypothesis, $H_1$: Some association
- We can explore these data at:-
    + http://bioinformatics.cruk.cam.ac.uk/apps/stats//contingency-table/

## Chi-square test: calculating expected frequencies

```{r}
df2 <- cbind(df, rowSums(df))
df2 <- rbind(df2, colSums(df))
colnames(df2)[3] <- rownames(df2)[3] <- "Total"
print(ascii(df2),type="rest")
```

$$E = \frac{row.total \times col.total}{overall.total} $$

- e.g. for row 1, column 1
$$\frac{84}{124} \times \frac{68}{124} \times 124 = \frac{84\times68}{124} = 46.1$$

    
## Chi-square test: calculating the chi-square statistic

***Observed frequencies:***

```{r}
csq <- chisq.test(df,correct = FALSE)
print(ascii(df),type="rest")
```

***Expected frequencies:***

```{r}
print(ascii(round(csq$expected,2)),type="rest")
```

$$\chi^2_1 = \frac{(44-46.06)^2}{46.06} + \frac{(40-37.94)^2}{37.94} + \frac{(24-21.94)^2}{21.94} + \frac{(16-18.06)^2}{18.06} = 0.63$$

## Chi-square test

```{r}
library(ggplot2)
degf <- csq$parameter

xmax <- max(4,csq$statistic)
xs <- seq(0, xmax, length.out = 10000)
df <- data.frame(X = xs, Y = dchisq(xs,degf))

title <- substitute(paste(chi^2, " with ", degf, " degrees of freedom"),list(degf=degf))

p <- ggplot(df, aes(x=X,y=Y)) + geom_line() + geom_vline(xintercept=csq$statistic,col="red") + xlim(0,xmax+1) + ggtitle(title)
p
pval <- round(csq$p.value,2)

```

Test statistic: ${\chi_1}^2$ = `r pval`
df = `r degf`
P-value = `r pval`

***Do not reject $H_0$ (No evidence of an association between treatment group and tumour shrinkage)***

## Limitations of the chi-square test

- In general, a Chi-square test is appropriate when:
    + at least 80% of the cells have an expected frequency of 5 or greater
    + none of the cells have an expected frequency less than 1 
- If these conditions aren’t met, ***Fisher’s exact test*** should be used.
    + no statistic is computed; an exact probability is calculated

## Same question, smaller sample size

- e.g. Research question: Is there an association between treatment group and tumour shrinkage?


```{r}
df <- data.frame(Tumour.Did.Not.Shrink = c(8,9), Tumour.Did.Shrink = c(3,4))
rownames(df) <- c("Treatment","Placebo")
df2 <- cbind(df, rowSums(df))
df2 <- rbind(df2, colSums(df))
colnames(df2)[3] <- rownames(df2)[3] <- "Total"
print(ascii(df2),type="rest")
```

- Null hypothesis: $H_0$: No association
- Alternative hypothesis: $H_1$: Some association

## Fishers' exact test; results

Observed frequencies:- 

```{r}
exp <- round(chisq.test(df)$expected,1)
print(ascii(df2),type="rest")
```

Expected frequencies:-

```{r}
print(ascii(exp),type="rest")
pv <- fisher.test(df)$p.value
```

- Test statistic: N/A
- P-value `r pv`
- Interpretation: ***Do not reject*** $H_0$ (No evidence of an association between treatment group and tumour shrinkage)

## Chi-square test for trend

- e.g. Research question: Is there a linear association between tumour grade and the incidence of tumour shrinkage

```{r}
df <- data.frame(Tumour.Did.Not.Shrink = c(18,15,11), Tumour.Did.Shrink = c(5,14,21))
rownames(df) <- c("Grade 1","Grade 2", "Grade 3")
df2 <- cbind(df, rowSums(df))
df2 <- rbind(df2, colSums(df))
colnames(df2)[3] <- rownames(df2)[4] <- "Total"
print(ascii(df2),type="rest")
```

- Null hypothesis: $H_0$: No linear association
- Alternative hypothesis: $H_1$: Some linear association

## Expected frequencies

```{r}
print(ascii(df2),type="rest")
```

Expected

```{r}
exp <- round(chisq.test(df)$expected,1)
exp
```

## CHi-square test for trend

```{r}
df <- data.frame(Tumour.Did.Not.Shrink = c(18,15,11), Tumour.Did.Shrink = c(5,14,21))
rownames(df) <- c("Grade 1","Grade 2", "Grade 3")
obs.prop <- df$Tumour.Did.Shrink/ rowSums(df)
plot(c(2,3,4),obs.prop,pch="x",col="blue",xlim=c(1,5),ylim=c(0,1),xlab="Tumour Grade",ylab="Proportion with tumour shrinkage ")
lm.h0 <- lm(obs.prop ~ c(2,3,4))
abline(lm.h0,col="blue")
abline(h=0.5,col="red")
points(c(2,3,4),rep(0.5,3),pch="x",col="red")
legend("topright", col=c("blue","red"),legend=c("H0","H1"),lty=c(1,1))
```


## Chi-square test for trend - Results

Test statistic: $$\chi_1^2 = 10.18$$
df = 1
P-value = 0.001

Conclusion:- Reject $H`_0$ (evidence of a linear association between tumour grade and tumour shrinkage) 

```{r fig.height=6,fig.width=6}
degf <- 1

xmax <- max(4,csq$statistic)
xs <- seq(0, xmax, length.out = 10000)
df <- data.frame(X = xs, Y = dchisq(xs,degf))

title <- substitute(paste(chi^2, " with ", degf, " degrees of freedom"),list(degf=degf))

p <- ggplot(df, aes(x=X,y=Y)) + geom_line() + ggtitle(title)
p
```


## Summary - Categorical variables

- Chi-square test
    + Use when we have two categorical variables, each with two or more levels, and our expected frequencies are not too small. 

- Fishers exact test
    + Use when we have two categorical variables, each with two levels, and our expected frequencies are small. 
 
- ***Exercise***
    + Complete "Tests for categorical variables" section of practical


## Small group Exercise

- In groups of 3 /4 you are going to explore several example datasets
- Each dataset has a brief explanation
    + You have to decide what test is appropriate?
    + Parametric / Non-parametric?
    + Paired? Independent?
    + Two-sided?
- Record your observations here:-
https://public.etherpad-mozilla.org/p/2016-11-28-intro-to-stats



## References

- Essential Medical Statistics, Betty Kirkwood and Jonathan Sterne, Wiley-Blackwell, 2nd Edition 2003.
- Practical Statistics for Medical Research, Douglas G. Altman, Chapman & Hall / CRC, 1999.    
- Ten Simple Rules for Effective Statistical Practice, Robert E. Kass, Brian S. Caffo, Marie Davidian, Xio-Li Meng, Bin Yu, Nancy Reid http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004961                

## Finally

- Course Materials:-

    + http://tiny.cc/crukStats

- Course Feedback:-

    + http://tinyurl.com/stats-nov18


